Race Science in Mainstream Psychology, 1960-2010

### 2025-04-09 paper/paper.qmd
giant caption

![*[caption]*](img/15_presence.png){#fig-presence}

@@

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (@fig-focal). @fig-presence shows the results of this analysis.

@fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

@fig-presence C shows that a greater share of papers in topic 24 had Pioneer-funded authors, compared to the other two topics, throughout the scope of the study but especially after about 1980.

@fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B). Points above the dark diagonal line (length of the thin colored lines) indicate that, in that year-topic combination, Pioneer-funded authors' paper share was greater than their author share, and in this sense had a disproportionate "presence." For topic 7 (MQ-style race science), Pioneer-funded authors frequently had a small, disproportionately *negative* presence. While for topic 22 (mainstream intelligence) Pioneer-funded authors had a small or modest disproportionately positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

To contextualize this measure of "presence," for each topic we used a permutation-style approach, drawing random sets of 15 authors (who ever published in the topic) and calculating paper share and author share in the same way as for Pioneer-funded authors. @fig-presence E shows the resulting distribution of mean presence (difference between paper share and author share) across 1,000 such permutations, with vertical lines for the observed values for the Pioneer-funded authors. Compared to the permutations, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.

>>>>

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors (distinguished by family name) for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually fewer than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point represents one year. If Pioneer-funded authors had publishing profiles (y-axis) proportionate to their authorship profiles (x-axis), points would be close to the dark line $y = x$. Points above this line indicate that Pioneer-funded authors had a relatively greater publishing presence than would be expected given their authorship profile. Length of thin vertical lines indicates difference between paper share and author share. \
(E) Empirical density plots of mean difference between paper share and author share for 1,000 randomly drawn sets of 15 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

*[more details in supplement]*

@@

Because this was a post hoc analysis, we elected to examine "presence" only for the three focal topics for the medium vocabulary, $k=40$ model (fig. @fig-focal). Fig. @fig-presence shows the results of this analysis.

Fig. @fig-presence A shows that topics 7 (MQ-style race science) and 24 (race-and-intelligence) had stable numbers of authors over the scope of the study, while topic 22 (mainstream intelligence) saw an exponential-like increase. Fig. @fig-presence B indicates that Pioneer-funded authors were a relatively high share of authors in topic 24, often comprising more than 10% of all authors in this topic after 1980. In contrast, these authors were almost never more than 10% of all authors in the other two topics.

Fig. @fig-presence C shows that a greater share of papers in topic 24 had Pioneer-funded authors, compared to the other two topics, throughout the scope of the study but especially after about 1980.

Fig. @fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B). Points above the dark diagonal line (length of the thin colored lines) indicate that, in that year-topic combination, Pioneer-funded authors' paper share was greater than their author share, and in this sense had a disproportionate "presence." For topic 7 (MQ-style race science), Pioneer-funded authors frequently had a small, disproportionately *negative* presence. While for topic 22 (mainstream intelligence) Pioneer-funded authors had a small or modest disproportionately positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

To contextualize this measure of "presence," for each topic we used a permutation-style approach, drawing random sets of 15 authors (who ever published in the topic) and calculating paper share and author share in the same way as for Pioneer-funded authors. Fig. @fig-presence E shows the resulting distribution of mean presence (difference between paper share and author share) across 1,000 such permutations, with vertical lines for the observed values for the Pioneer-funded authors. Compared to the permutations, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.
